Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-06-01 article/main.tex
ack
sections 3, 4

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{minted}
\usepackage{epigraph}
\usepackage{cellspace}
\usepackage{multirow}
%\usepackage{parskip}
\usetikzlibrary{patterns}

\DeclareMathSymbol{:}{\mathord}{operators}{"3A}
\DeclareMathOperator{\LCG}{LCG}
\DeclareMathOperator{\MCG}{MCG}
\DeclareMathOperator{\MAJ}{MAJ}

\usepackage[noend]{algpseudocode}

@@

\todo{Mentionner Knuth et les LCG + Frieze + Boyar + Stern...}

@@

This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
has an internal state of 128-bit, which operate as a linear congruential
generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]

Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

Each time the PRNG is clocked, 64 output bits are extracted from the internal
state using a non-linear function that makes use of data-dependent rotations, in
a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
significant bits of the internal state encode a number $0 \leq r < 64$. The two
64-bit halves of the internal state are XORed together, and this 64-bit result
is rotated right by $r$ positions. The successive 64-bit outputs of the
generator are $X_0, X_1, \dots$ where:
\begin{equation}\label{eq:output}
  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
\end{equation}

Figure~\ref{pcg128out} summaries the process. The overall design strategy is
similar to that of a filtered LFSR: the successive states of a weak internal
generator with a strong algebraic structure are ``filtered'' by a non-linear
function.

Updating the internal state requires a $128 \times 128 \rightarrow 128$
multiplication operation. In fact, this can be done with two
$64 \times 64 \rightarrow 64$ multiplications, one
$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
desktop CPUs all implement these operations in hardware, so the generator is
relatively fast on these platforms.

@@

\section{Reconstructing Truncated Geometric progressions}
\label{sec:geometric}

This is a recurring theme is this paper, so we now state the problem in
general. If $a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there
is a \emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
$0 \leq 1 \leq \beta$) such that:
\begin{align}
  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
\end{align}

\subsection{Reconstructing Linear Congruential Generators}

Given an integer \(k\), a fixed multiplier \(a\) and an increment \(c\), define the
sequence \[u_0 = x, u_{i+1} = a u_i + c \bmod 2^k\] The sequence \(u\) form the
successive internal states of a linear congruential generator (LCG).

Let $\LCG_{k}(x, c)$ denote the vector $(u_0, u_1, u_2, \dots)$ of integers
modulo $2^k$.  It is easy to check that:
\begin{align}
\label{eq:lcg-additive}
\LCG_{k}(x + y, c + d) &= \LCG_{k}(x, c) + \LCG_{k}(y, d),  \\
\label{eq:lcg-scalar}
\LCG_{k}(\lambda x, \lambda c) &= \lambda \LCG_{k}(x, c).   
\end{align}
Hence those sets of vector have a lot of structure we can use.

A reconstruction algorithm has been proposed by D. Knuth in 1985\cite{Knuth85}. The LCG he considered revealed only the upper bits of each internal state. The modulus \(2^k\) was known while the multiplier \(a\) and the increment \(c\) weren't. This algorithm could reconstruct the LCG in \(\mathcal{O}(2^l)\) where \(l\) is the number of non revealed bit of each internal state.

In 1989, J. Boyar improve the result in a first way \cite{Boyar1989}. He presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.

In 1988, A. Frieze \textit{et al} \cite{Frieze} improve the result in a second way. They supposed that the multiplier \(a\) and the modulus \(2^k\) were known and used lattice-based technique to recover the LCG more efficiently.

Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment \(c\) is zero) \cite{Lehmer}.

\subsection{The lattice associated to a geometric progression}\label{lattice}

We fix a multiplier \(a\), integers \(k\), \(n\) and we consider \(L(n,k)\) the set of \(\LCG_{n,k}(x, 0)\), for \(x\) in \(\mathbb{Z}_{2^k}\). Then

\begin{itemize}
	\item By \eqref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
	\item By \eqref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
\end{itemize}

Hence \(L(n,k)\) is a lattice. We easily notice that \(L(n,k)\) is spawned by the lines of the following \(n\times n\) matrix :

\[
  \begin{pmatrix}
    1 & a   & a^2 & \dots & a^{n-1} \\
    0 & 2^k & 0   & \dots & 0 \\
    0 & 0   & 2^k & \dots & 0 \\
    \dots & \dots & \dots & \dots & \dots\\
    0 & 0 & 0 & \dots & 2^k \\
  \end{pmatrix}
\]

Once again, we will omit $n$ and $k$ when they are obvious from the context.

We want to find an element \(U\) in the lattice only knowing the $\ell$ most significant bits of each of its coordinates.  More precisely, let $T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$. We assume that there is a ``noise vector'' $N$ such that $N_i \in \{-1, 0, 1\}$, and we set $\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$. The problem we have is the  following: recover $U$ from its noisy truncation $\widetilde{T}$.

In principle, $2^{k-\ell} \widetilde{T}$ should be ``close'' to \(U\). We must distinguish three cases.
\begin{itemize}
	\item If for each \(i\) the integer \(\widetilde{T}_i\) satisfies \(1\leqslant\widetilde{T}_i\leqslant 2^{k-l}-2\) then \(\widetilde{T}_i = T_i + N_i\) (without modulo). So

	\begin{align*}
	\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert &=\lvert 2^{k-\ell}(T_i + N_i)-U_i \rvert \\
	 &= \lvert 2^{k-\ell}N_i - U_i[0,k-\ell] \rvert\\
	 &\leqslant \lvert 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \rvert \\
	 &\leqslant 2^{k-\ell} + 2^{k-\ell}-1\\
	 &\leqslant 2^{k-\ell +1 }- 1
	\end{align*}

	\item If there are indexes such that \(\widetilde{T}_i = 2^\ell-1\).
	\begin{itemize}
		\item If \(N_i \in \{0,1\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
		\item If \(N_i = -1\), then \(U_i=U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \rvert =\lvert 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \rvert = \lvert 2^{k-\ell} + U_i[0,k-\ell]\rvert \leqslant 2^{k-\ell + 1} -1 \]
	\end{itemize}

	\item If there are indexes such that \(\widetilde{T}_i = 0\).
	\begin{itemize}
		\item  If  \(N_i \in \{-1,0\}\), then \(\widetilde{T}_i = T_i + N_i\) (without modulo) and \[\lvert 2^{k-\ell}\widetilde{T}_i-U_i \rvert \leqslant \leqslant 2^{k-\ell +1 }- 1\]
		\item If \(N_i = 1\) , then \(U_i= 2^k-2^(k-l) + U_i[0,k-\ell]\) and  \[\lvert 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \rvert =\lvert 0 -(U_i - 2^k) \rvert = \lvert 2^k-2^(k-l) + U_i[0,k-\ell] - 2^k\rvert \leqslant 2^{k-\ell + 1} -1 \]
	\end{itemize}
\end{itemize}
	We need a new vector \(U'\) defined as follow.
\begin{itemize}
	\item If \(T_i = 0\) and \(N_i = -1\), \(U'_i=U_i+2^k\).
	\item If \(T_i = 2^\ell - 1\) and \(N_i = +1\), \(U'_i=U_i-2^k\).
	\item Else \(U'_i=U_i\).
\end{itemize}
In all this cases \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \), and \(U' \equiv U \mod 2^k \).

\todo{C'est un lemme mais pour l'instant y'a pas de commande lemme}
\begin{theorem}
	If $\widetilde{T}$ is the noisy truncation of \(U\) of \(\ell\) bits, then there exists \(U'\) such that :
	\begin{itemize}
		\item \(U'\equiv U \mod 2^k\)
		\item \(\lVert U' -2^{k-\ell} \widetilde{T} \rVert \leqslant \sqrt{n}(2^{k-\ell+1}-1) \)
	\end{itemize}
\end{theorem}

\subsection{A first way to find U : using an exact CVP solver}

Solving exactly the CVP is a hard computational problem. The algorithms used are costly, but we can afford to use them up to dimension 50.

We define \(U'\) as in subsection \ref{lattice}. We want to know if \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed \(U'\).

What can we say about \(|CVP(L,2^{k-\ell}\widetilde{T})-U'|\) ?

\begin{align*}
|CVP(L,T)-U| &= |CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
& \leqslant |CVP(L,T)-2^{k-\ell}\widetilde{T}|+|2^{k-\ell}\widetilde{T}-U'|\\
\end{align*}

But, as the vector \(U'\) is in the lattice and by definition of the CVP, \(|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leqslant |2^{k-\ell}\widetilde{T}-U'| \), hence 
\[|CVP(L,2^{k-\ell}\widetilde{T})-U'|\leqslant 2|2^{k-\ell}\widetilde{T}-U'|\]

If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,2^{k-\ell}\widetilde{T})\) is indeed the vector we were searching for.

To compute \(\lambda_1(L)\), we find the shortest vector of L thanks to an exact SVP solver (SVP stands for Shortest Vector Problem) and we compute its norm.

In this article, we will only use an exact CVP solver when \(k = 128\) and \(\ell = 6\). So, if we fix \(\ell = 6\) we can search for the minimum \(n\) such that \[2\sqrt{n}(2^{128-6+1}-1) \leqslant \lambda_1(L(128,n))\].

After computation, we find that the minimal \(n\) is 63.

\subsection{A second way to find U : the Babai's rounding}

As using an exact CVP solver is costly, we will try a cheaper but less precise way to find the closest vector.

We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \sup_{x \in \mathbb{R}^n}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert \cdot \rVert_2\) norm,
\(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular
value of $G$ (equivalently, $\sigma(G)$ is the absolute value of the largest
eigenvalue of $G^{t} G$).

As \(2^{k-\ell}\widetilde{T}\) is not an element of the lattice, \(2^{k-\ell}\widetilde{T}G^{-1}\) is not an integer vector.

We fix \(r = rounding(2^{k-\ell}\widetilde{T}G^{-1}) \), then \(rG\) is an element of the lattice. What can we say about the distance between \(rG\) and \(U'\) ?

\begin{align*}
\lVert U' - rG \rVert &= \lVert U - (r-2^{k-\ell}\widetilde{T}G^{-1} + 2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
&= \lVert U' - 2^{k-\ell}\widetilde{T} + (r-2^{k-\ell}\widetilde{T}G^{-1})G \rVert\\
&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
\end{align*}

By definition, \(r\) is the closest integer vector to \(2^{k-\ell}\widetilde{T}G^{-1}\). But as \(U'\) is an element of the lattice, \(U'G^{-1}\) is an integer vector. Thus \(r-2^{k-\ell}\widetilde{T}G^{-1}\) is shorter than \(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1}\).

Hence :
\begin{align*}
\lVert U' - rG \rVert &\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(r-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'G^{-1}-2^{k-\ell}\widetilde{T}G^{-1})\rVert \times \lVert G\rVert\\	
&\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert + \lVert(U'-2^{k-\ell}\widetilde{T})\rVert \times \lVert G^{-1} \rVert  \times \lVert G\rVert\\
& 	\leqslant \lVert U' - 2^{k-\ell}\widetilde{T} \rVert \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert )\\
\end{align*}

If we can prove that the right side of the inequality is smaller than the minimum of the lattice \(\lambda_1(L)\), then we have proved that \(rG\) is indeed the vector we were searching for.

In this article, we will only use the Babai's rounding when \(k = 64\). Hence \(\lVert 2^{k-\ell}\widetilde{T}-U' \rVert \leqslant \sqrt{n}2^{64-\ell+1}-1 \). So, if we fix \(n\) we can search for the minimum \(\ell\)\todo{ATTENTION ici \(\ell\) c'est le nombre de bit devinés, donc l'ancien \(6+\ell\)} such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{64-\ell+1}-1 \leqslant \lambda_1(L(n,64))\]

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amssymb, amsmath, amsfonts, amscd}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xspace}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{tikz}
\usepackage{minted}
\usepackage{epigraph}
\usepackage{cellspace}
\usepackage{multirow}
\usetikzlibrary{patterns}
\usepackage[noend]{algpseudocode}

\DeclareMathSymbol{:}{\mathord}{operators}{"3A}
\DeclareMathOperator{\LCG}{LCG}
\DeclareMathOperator{\MCG}{MCG}
\DeclareMathOperator{\MAJ}{MAJ}
\DeclareMathOperator{\CVP}{CVP}

% horrible hack
% Math symbol font matha
\DeclareFontFamily{U}{matha}{\hyphenchar\font45}
\DeclareFontShape{U}{matha}{m}{n}{
      <5> <6> <7> <8> <9> <10> gen * matha
      <10.95> matha10 <12> <14.4> <17.28> <20.74> <24.88> matha12
      }{}
\DeclareSymbolFont{matha}{U}{matha}{m}{n}
\DeclareFontSubstitution{U}{matha}{m}{n}

% Math symbol font mathb
\DeclareFontFamily{U}{mathx}{\hyphenchar\font45}
\DeclareFontShape{U}{mathx}{m}{n}{
      <5> <6> <7> <8> <9> <10>
      <10.95> <12> <14.4> <17.28> <20.74> <24.88>
      mathx10
      }{}
\DeclareSymbolFont{mathx}{U}{mathx}{m}{n}
\DeclareFontSubstitution{U}{mathx}{m}{n}

% Symbol definition
\DeclareMathDelimiter{\vvvert}{0}{matha}{"7E}{mathx}{"17}

@@

A reconstruction algorithm has been proposed by D. Knuth in 1985\cite{Knuth85}. The LCG he considered revealed only the upper bits of each internal state. The modulus $2^k$ was known while the multiplier $a$ and the increment $c$ weren't. This algorithm could reconstruct the LCG in $\mathcal{O}(2^l)$ where $l$ is the number of non revealed bit of each internal state.

In 1989, J. Boyar improve the result in a first way \cite{Boyar1989}. She presented an algorithm which could predict a LCG where all the parameters (multiplier, increment and initial state) where unknown under the condition that the number of bit unrevealed was really small in comparison to the modulus.

In 1988, A. Frieze \textit{et al} \cite{Frieze} improve the result in a second way. They supposed that the multiplier $a$ and the modulus $2^k$ were known and used lattice-based technique to recover the LCG more efficiently.

Here we will use techniques presented in \cite{Frieze} to reconstruct some Lehmer's generator (LCG where the increment $c$ is zero) \cite{Lehmer}.

@@

This section introduces some notations and describes the \textsf{PCG64}
non-cryptographic pseudo-random number generator (a.k.a. \textsf{PCG-XSL-RR} in
the designer's terminology).

If $x \in \{0, 1\}^n$ is an $n$-bit string, then $x[i:j]$ denotes the bit string
$x_ix_{i+1}\dots x_{j-2}x_{j-1}$, where $x = x_0\dots x_{n-1}$ (this is the
``slice notation'' used in \textsf{Python}). The set $\mathbb{Z}_{2^k}$ of
integers modulo $2^k$ is seen as the set of $k$-bit strings. If $x$ is a
floating point number, then $\lfloor x \rceil$ denotes the nearest integer
(using the ``rounding half to even'' tie-breaking rule --- this is the default
in IEEE754 arithmetic). If $U$ is a vector or a sequence, then $U_i$ is the
$i$-th element (we use capital letters for these). If $U$ is such a sequence, we
denote by $U \bmod M$ the sequence $(U_0 \bmod M, U_1 \bmod M, \dots)$. The XOR
operation is denoted $\oplus$, left and right rotations are denote $\lll$ and
$\ggg$ respectively. Modular addition is denoted $+$ (or $\boxplus$ to make it
even more explicit).

\textsf{PCG64} has an internal state of 128-bit, which operate as a linear
congruential generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where is a fixed 126-bit constant. The first initial state $S_0$ is the seed of
the generator. The increment $c$ may be specified by the user of the PRNG to
produce different output streams with the same seed (just as the IV acts in a
stream cipher). If no value of $c$ is specified, then a default increment is
provided. Note that $c$ must be odd. The default values are:
\begin{align*}
  a &= \phantom{1}47026247687942121848144207491837523525 & \text{(fixed)}\\
  c &= 117397592171526113268558934119004209487           & \text{(default value, user-definable)}
\end{align*}

Each time the PRNG is clocked, 64 output bits are extracted from the internal
state using a non-linear function that makes use of data-dependent rotations, in
a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
significant bits of the internal state encode a number $0 \leq r < 64$. The two
64-bit halves of the internal state are XORed together, and this 64-bit result
is rotated right by $r$ positions.

The successive 64-bit outputs of the generator are $X_0, X_1, \dots$ where:
\begin{equation}\label{eq:output}
  X_i =(\underbrace{S_i[0:64] \oplus S_i[64:128]}_{Y_i}) \ggg \underbrace{S_i[122:128]}_{r_i}.
\end{equation}
For the sake of convenience, we denote by $Y_i$ the XOR of the two halves of the
state (before the rotation) and by $r_i$ the number of shifts of the ``$i$-th
rotation''.

Fig.~\ref{pcg128out} summaries the process. The overall design strategy is
similar to that of a filtered LFSR: the successive states of a weak internal
generator with a strong algebraic structure are ``filtered'' by a non-linear
function.

Updating the internal state requires a $128 \times 128 \rightarrow 128$
multiplication operation. In fact, this can be done with three
$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
desktop CPUs all implement these operations in hardware, so the generator is
quite fast on these platforms.

@@

\section{Tools}
\label{sec:geometric}

In the rest of this paper, we often perform arithmetic operations on integers
where only some bits are known. This leads to generation of unknown carries. If
$a, b$ are integers modulo $2^{128}$ and $0\leq i<j<128$, then there is a
\emph{carry} $0 \leq \gamma \leq 1$ (resp. a \emph{borrow}
$0 \leq 1 \leq \beta$) such that:
\begin{align}
  (a \boxplus b)[i:j]  &= a[i:j] \boxplus  b[i:j] \boxplus  \gamma, \label{eq:caryy}  \\
  (a \boxminus b)[i:j] &= a[i:j] \boxminus b[i:j] \boxminus \beta.  \label{eq:borrow}
\end{align}

\subsection{Linear Congruential Generators and Lattices}

Given an integer $k$, a fixed multiplier $a$, an increment $c$ and a ``seed''
$x$, define the sequence:
\[
  U_0 = x, U_{i+1} = a U_i + c.
\]
When reduced modulo $2^k$, the sequence $U$ form the successive states of a
linear congruential generator (LCG). Let $\LCG_{k}(x, c)$ denote the vector
$(u_0, u_1, u_2, \dots)$ of integers modulo $2^k$. It is easy to check that:
\begin{align}
\label{eq:lcg-additive}
\LCG_{k}(x + y, c + d) &= \LCG_{k}(x, c) + \LCG_{k}(y, d),  \\
\label{eq:lcg-scalar}
\LCG_{k}(\lambda x, \lambda c) &= \lambda \LCG_{k}(x, c).   
\end{align}

Let $\mathcal{L}$ denote the euclidean lattice spawned by the rows of the
following $n\times n$ matrix:
\[
  G_{n,k} = \begin{pmatrix}
    1 & a   & a^2 & \dots & a^{n-1} \\
    0 & 2^k & 0   & \dots & 0 \\
    0 & 0   & 2^k & \dots & 0 \\
    \dots & \dots & \dots & \dots & \dots\\
    0 & 0 & 0 & \dots & 2^k \\
  \end{pmatrix}
\]
This lattice contains all $n$-terms geometric progressions of common ratio $a$
modulo $2^k$; therefore the first $n$ terms of the sequence $\LCG_{k}(x, 0)$
give the coordinates of a point of this lattice.

Reconstructing the state of a truncated linear congruential generator can
generally be seen as the problem of finding a point of this lattice given only
an approximation thereof, for instance when the least-significant bits of each
components have been dropped. The ``lattice approach'' to truncated linear
congruential generators is due to Frieze \textit{et al.}~\cite{Frieze}.

Let $U$ be a vector of $n$ integers modulo $2^k$ such that
$U_{i+1} = a U_i \bmod 2^k$; it is a geometric progression of common ratio $a$,
modulo $2^k$, therefore $(U_0, \dots, U_{n-1}) \in \mathcal{L}$. Let
$T_i = U_i[k-\ell:k]$ denote the top $\ell$ bits of $U_i$, and let $N$ denote an
arbitrary ``noise vector'' such that $N_i \in \{-1, 0, 1\}$. Finally, set
$\widetilde{T}_i = T_i + N_i \bmod 2^{\ell}$.  We will be facing the following
problem (``reconstructing noisy truncated geometric series'') several times:
\begin{description}
\item[INPUT]  $\widetilde{T} = (\widetilde{T}_0, \dots, \widetilde{T}_{n-1}) \in \left(\mathbb{Z}_{2^\ell}\right)^n$, a (noisy) version of $U$ truncated to the top $\ell$ bits.
\item[OUTPUT] $U_0 \bmod 2^k$, the first term of the (non-truncated) geometric sequence.
\end{description}

We will be facing a ``high-dimension'' instance in section~\ref{sec:big_cvp}
and many ``low-dimension'' in sections~\ref{sec:Cknown}
and~\ref{sec:Cunknown}. The rest of this section discusses algorithmic tools to
solve these problems. We first claim that $2^{k-\ell} \widetilde{T}$ is
``close'' to a point of the lattice $\mathcal{L}_{n,k}$.

\begin{lemma}\label{thm:close}
  There exists $U' \in \mathcal{L}$ such that $U'_i \equiv U_i \bmod 2^k$
  and $\left\lVert U' -2^{k-\ell} \widetilde{T} \right\rVert \leq 2 \sqrt{n} 2^{k-\ell}$.
\end{lemma}

\begin{proof}
  We first observe that $U$ belongs to the lattice $\mathcal{L}$. We start by setting $U' \gets U$, and we examine all coordinates of $U$:
  \begin{itemize}
  \item If $\widetilde{T}_i = T_i + N_i$ (without modulo), then we have:
    \begin{align*}
      \left| 2^{k-\ell}\widetilde{T}_i-U_i \right| &= \left| 2^{k-\ell}(T_i + N_i)-U_i \right| 
                                                     = \left| 2^{k-\ell}N_i - U_i[0,k-\ell] \right| \\
                                                   &\leq \left| 2^{k-\ell}N_i\rvert + \lvert U_i[0,k-\ell] \right| 
                                                   % &\leq 2^{k-\ell} + 2^{k-\ell} - 1\\
      \leq 2^{k-\ell +1 }.
    \end{align*}

  \item Otherwise, there are two possible ``wraparound'' cases:
    \begin{itemize}
    \item Either $T_i = 0$ and $N_i = -1$, which leads to $\widetilde{T}_i = 2^\ell-1$. 
      In this case, we have $U_i = U_i[0,k-\ell]$ and we set $U'_i = U_i + 2^{k}$ (note that this amount to adding a lattice vector to $U'$, so $U'$ stays in the lattice). We have:
      \[
        \left| 2^{k-\ell}\widetilde{T}_i-(U_i + 2^k) \right| = %\left| 2^{k-\ell}(2^\ell -1) -(U_i + 2^k) \right| =
        \left| 2^{k-\ell} + U_i[0,k-\ell]\right| \leq 2^{k-\ell + 1}. 
      \]

    \item Or $T_i = 2^{\ell}-1$ and $N_i = +1$, which leads to
      $\widetilde{T}_i = 0$.  This implies that
      $U_i= 2^k-2^{k-\ell} + U_i[0,k-\ell]$; we set $U'_i = U_i - 2^k$ (again,
      with this modification $U'$ stays in the lattice), and we find:
      \[
        \left| 2^{k-\ell}\widetilde{T}_i-(U_i - 2^k) \right| %=\left| 0 -(U_i - 2^k) \right|
        = \left| 2^k-2^{k-\ell} + U_i[0,k-\ell] - 2^k\right| \leq 2^{k-\ell + 1}.
      \]
    \end{itemize}
  \end{itemize}
  In the end, we have
  $\left\| U' -2^{k-\ell} \widetilde{T} \right\| \leq 2 \sqrt{n} 2^{k-\ell}$,
  and $U' \equiv U \mod 2^k $.
\end{proof}

\subsection{Reconstruction in ``High'' Dimension Using an Exact CVP Solver}

Lemma~\ref{thm:close} tells us that the approximation of a geometric sequence
obtained by dropping least-significant bits cannot be arbitrarily far from a
lattice point which reveals $U_0 \bmod 2^k$. Therefore, we may possibly
reconstruct truncated geometric series by finding the lattice vector closest to
the approximation we have. This means solving instances of the well-known
\textsf{Closest Vector Problem} (CVP), a fundamental hard problem on
lattices. It is NP-hard, and all known algorithms are exponential in the
dimension of the lattice, yet they can be fairly practical up to dimension
$\approx 70$.

Let $\CVP(\mathcal{L}, x)$ denote the vector of $\mathcal{L}$ closest to the
input vector $x$. Using the same notations as above, we want to know if
$\CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T}\right)$ is indeed $U'$. This
will necessarily be the case when $\|2^{k-\ell}\widetilde{T} - U'\|$ is smaller
than the length of the shortest non-zero vector of $\mathcal{L}$ --- this quantity,
the \emph{first minimum} of the lattice, is denoted by $\lambda_1(\mathcal{L})$.
%What can we say about $|CVP(L,2^{k-\ell}\widetilde{T})-U'|$ ?
By the triangular inequality, we have:
\begin{align*}
|\CVP(\mathcal{L}, T) - U'| %&= \left| \CVP\left( \mathcal{L}, 2^{k-\ell}\widetilde{T} \right) \riht| + \left|2^{k-\ell}\widetilde{T} - U'\right| %-2^{k-\ell}\widetilde{T}+2^{k-\ell}\widetilde{T}-U'|\\
                             & \leq \left| \CVP\left(\mathcal{L}, 2^{k-\ell}\widetilde{T} \right) - 2^{k-\ell}\widetilde{T}\right | + \left|2^{k-\ell}\widetilde{T} - U'\right|.
\end{align*}
But, as the vector $U'$ belongs to the lattice, by definition of the closest
vector and by lemma~\ref{thm:close}:
% $|CVP(L,2^{k-\ell}\widetilde{T})-2^{k-\ell}\widetilde{T}| \leq
% |2^{k-\ell}\widetilde{T}-U'| $, hence
\[
  |\CVP(\mathcal{L}, T) - U'| \leq 2 \left|2^{k-\ell} \widetilde{T} - U' \right| \leq 4 \sqrt{n} 2^{k-\ell}.
\]

If we can prove that the right side of this inequality is smaller than the first
minimum of the lattice $\lambda_1(\mathcal{L})$, then we would have proved that
$\CVP(\mathcal{L}, T)$ indeed reveals $U_0 \bmod 2^k$.

In section~\ref{sec:big_cvp}, we will be facing the problem of reconstructing a
geometric sequence modulo $2^{128}$ given arbitrarily many (noisy versions of
the) most-significant $6$ bits of successive elements of the sequence. Therefore
we have $k=128$ and $\ell=6$, and we wish to determine the required number of
samples, \textit{i.e.} the value of $n$. This means finding the values of $n$
such that $\sqrt{n} 2^{124} \leq \lambda_1\left(\mathcal{L}\right)$.

Starting from $n=\lceil 122/6\rceil\rceil$, we computed the length of the
shortest vector of the lattice spanned by $G_{n, 128}$ for each successive $n$
until the condition holds. The \textsf{Shortest Vector Problem} is another
well-known lattice NP-hard problem; we used the (almost) off-the-shelf
\texttt{G6K} library~\cite{AlbrechtDHKPS19}, which gave results very quickly by
sieving. \texttt{fplll}~\cite{fplll} was too slow above dimension 50, in the
default settings.

After this computation, we found that the minimal possible $n$ is 63: with
$n=63$, the shortest vector of $\mathcal{L}$ has length greater than
$2^{127.02}$, which is high enough. This vector can be obtained by bootstrapping
the geometric sequence with
\[
  U_0 = 12144252875850345479015002205241987363
\]
then reducing the terms modulo $2^{128}$ in zero-centered representation
(subtracting $2^{128}$ to $U_i$ if $U_i > 2^{127}$). It follows that when
$n\geq 63, k=128$ and $\ell=6$, any CVP oracle will return a vector congruent to
the original $U$ when given $\widetilde{T}$.

\subsection{Reconstruction in Low Dimension Using Babai's Rounding}

In sections~\ref{sec:Cknown} and~\ref{sec:cvp_small_2} we will need to
reconstruct billions of noisy truncated geometric series modulo $2^{64}$ with
very few terms, of which a large fraction of most-significant bits are known. In
this setting, the CVP problem becomes much easier. This enables us to use faster
and more \textit{ad hoc} methods, such as Babai's rounding
algorithm~\cite{Babai86}.

If $M$ is a square matrix, we denote by $\vvvert M \vvvert$ the induced matrix norm :
\[
  \vvvert M \vvvert =  \sup_{x \in \mathbb{R}^n} \frac{\| xM \|}{\| x \|}
\]
In the case of the $\| \cdot \|_2$ norm used throughout this paper, $\| M \|$ is
the largest singular value of $G$; equivalently, it is the square root of the
absolute value of the largest eigenvalue of $M^{t} M$.
